---
title: "Retail Strategy Analytics"
output:
  pdf_document:
    df_print: default
    highlight: tango
    keep_tex: yes
    latex_engine: xelatex
header-includes:
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  - \usepackage{fontspec}
  - \setmainfont{Roboto}
  - \setmonofont{Consolas}
---

```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(linewidth=80)
```
```{r knitr line wrap setup, include=FALSE}
# set up line wrapping in MD knit output
library(knitr)
hook_output = knit_hooks$get("output")
knit_hooks$set(output = function(x, options)
{
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth))
    {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n))
      x = strwrap(x, width = n)
    x = paste(x, collapse = "\n")
    }
  hook_output(x, options)
})
```

## Load required libraries and datasets
```{r 0 Load Libraries, results = 'hide'}
#### Example code to install packages
#install.packages("data.table")
#### Load required libraries
library(tidyverse)
library(data.table)
library(ggplot2)
library(readr)

filePath <- "C:/Users/U/Desktop/Quantium Analytics/"
transactionData <- fread(paste0(filePath,"QVI_transaction_data.csv"))
customerData <- fread(paste0(filePath,"QVI_purchase_behaviour.csv"))
```

## Exploratory data analysis
The first step in any analysis is to first understand the data. 
### Examining transaction data

```{r Examining transaction data}
#### Examine transaction data
str(transactionData)
```
We can see that the date column is in an integer format. Let's change this to a date format.
```{r Convert DATE to date format}
## A quick search online tells says that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```
Check that we are looking at the right products by examining PROD_NAME.
```{r Summary of PROD_NAME}
#### Examine PROD_NAME
summary(transactionData$PROD_NAME)
glimpse(transactionData$PROD_NAME)
```
Some basic text analysis by summarising the individual words in the product name.
```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries such as products that are not chips
productWords <- data.table(unlist(strsplit(unique(transactionData[, PROD_NAME]), "
")))
setnames(productWords, 'words')
```
Removing all words with digits and special characters such as '&' from our set of product words. 
```{r}

#### Defining the pattern
pattern <- "\\b\\w*[^a-zA-Z\\s]+\\w*\\b"

#### Removing digit and special characters
productWords$words <- sapply(productWords$words, function(text){
  words <- unlist(strsplit(text,"\\s+"))
  words <- words[!grepl(pattern,words)]
  paste(words, collapse = " ")
  })
###Most common words by counting the number of times a word appears and sorting them by this frequency in order of highest to lowest frequency

# combine all the cleaned text into a single string
all_text <- paste(productWords$words, collapse = " ")

# split the text into individual words
words <- unlist(strsplit(all_text, "\\s+"))

# create a frequency table
word_freq <- table(words)

# sort by frequency (descending order)
sorted_word_freq <- sort(word_freq, decreasing = TRUE)

# convert to a dataframe for easier viewing
sorted_word_freq_df <- as.data.frame(sorted_word_freq)


```
There are salsa products in the dataset but i am only interested in the chips category.
```{r}
#### Remove salsa products
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData <- transactionData[SALSA == FALSE, ][, SALSA := NULL]
```
Next, use `summary()` to check summary statistics such as mean, min and max values for each feature to see if there are any obvious outliers in the data and if there are any nulls in any of the columns (`NA's : number of nulls` will appear in the output if there are any nulls).
```{r initial summary}
#### Summarise the data to check for nulls and possible outliers
summary(transactionData)
```
There are no nulls in the columns but product quantity appears to have an outlier. Investigating further the case where 200 packets of chips are bought in one transaction.
```{r }
#### Filter the dataset to find the outlier
filter(transactionData, PROD_QTY == 200)
```
There are two transactions where 200 packets of chips are bought in one transaction and both of these transactions were by the same customer.
```{r}
#### see if the customer has had other transactions
filter(transactionData,LYLTY_CARD_NBR == 226000)
```
It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. removing this loyalty card number from further analysis.
```{r}
#### Filter out the customer based on the loyalty card number
transactionData <- transactionData %>% 
  filter(LYLTY_CARD_NBR != 226000)
#### Re-examine transaction data
summary(transactionData)
```
That's better. Looking at the number of transaction lines over time to see if there are any obvious data issues such as missing data.
```{r}
#### Count the number of transactions by date
transactionData %>% group_by(DATE) %>% 
  summarize(count = n())
```
There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.
```{r fig.align = "center"}
#### Create a sequence of dates and join this the count of transactions by date

# create sequence of date
date_sequence <- seq(as.Date("2018-07-01"),as.Date("2019-06-30"),by = "day")

# convert it to a data frame
date_df <- data.frame(DATE = date_sequence)

#perform a full join to merge the data with the date sequence
transactionData <- merge(date_df, transactionData, by = "DATE", all.x = TRUE)

# count number of transactions by day
transactions_by_day <- transactionData %>% 
  group_by(DATE) %>% 
  summarize(count = n())
#### Setting plot themes to format graphs
theme_set(theme_bw()) 
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(transactions_by_day, aes(x = DATE, y = count)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
see that there is an increase in purchases in December and a break in late December. zoom in on this.
```{r fig.align = "center"}
#### Filter to December and look at individual days
december_transactions = transactions_by_day %>% 
  filter(month(DATE) == 12)

theme_set(theme_bw()) 
theme_update(plot.title = element_text(hjust = 0.5))
#### Plot transactions over time
ggplot(december_transactions, aes(x = DATE, y = count)) +
  geom_line() +
  labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
  scale_x_date(breaks = "1 day") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on Christmas day.
Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.
```{r Create pack size}
#### Pack size
#### taking the digits that are in PROD_NAME
transactionData$PACK_SIZE <- parse_number(transactionData$PROD_NAME)

#### check if the pack sizes look sensible
summary(transactionData$PACK_SIZE)

```
The largest size is 380g and the smallest size is 70g - seems sensible!
```{r }
#### histogram of PACK_SIZE 
ggplot(transactionData, aes(x = PACK_SIZE)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") + 
  labs(title = "Histogram of Transactions by Pack Size", x = "Pack Size", y = "Number Of Transactions") + 
  theme_minimal()
```
Pack sizes created look reasonable.
Now to create brands, we can use the first word in PROD_NAME to work out the brand name...
```{r Create brand name}
#### Brands
transactionData$BRAND_NAME <- str_extract(transactionData$PROD_NAME, "^\\w+")
#### Checking brands
unique(transactionData$BRAND_NAME)
```
Some of the brand names look like they are of the same brands - such as RED and RRD, which are both Red Rock Deli chips. combining these together.
```{r Clean brand names}
#### Clean brand names
transactionData$BRAND_NAME <- replace(transactionData$BRAND_NAME, transactionData$BRAND_NAME == "Red","RRD")
transactionData$BRAND_NAME <- replace(transactionData$BRAND_NAME, transactionData$BRAND_NAME == "Smith","Smiths")
transactionData$BRAND_NAME <- replace(transactionData$BRAND_NAME, transactionData$BRAND_NAME == "Snbts","Sunbites")
transactionData$BRAND_NAME <- replace(transactionData$BRAND_NAME, transactionData$BRAND_NAME == "Dorito","Doritos")
transactionData$BRAND_NAME <- replace(transactionData$BRAND_NAME, transactionData$BRAND_NAME == "Infzns","Infuzions")
#### Check again
unique(transactionData$BRAND_NAME)
```
### Examining customer data
look at the customer dataset.
```{r 1 Exploratory data analysis}
#### Examining customer data
str(customerData)
summary(customerData)
```
```{r }
#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```
As the number of rows in `data` is the same as that of `transactionData`,no duplicates were created. `data` was created by setting `all.x = TRUE` (in other words, a left join) which means take all the rows in `transactionData` and find rows with matching values in shared columns and then joining the details in these rows to the `x` or the first mentioned table. check if some customers were not matched on by checking for nulls.
```{r Check for missing customer details}
sum(is.na(data$LYLTY_CARD_NBR))
```
Great, there are no nulls! So all customers in the transaction data has been accounted for in the customer dataset.

```{r Code to save dataset as a csv}
fwrite(data, paste0(filePath,"QVI_data.csv"))
```
Data exploration is now complete!
## Data analysis on customer segments
Now that the data is ready for analysis, we can define some metrics of interest:
- Who spends the most on chips (total sales), describing customers by lifestage and how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment
could also ask for more information. Examples are:
- The customer's total spend over the period and total spend for each transaction to understand what proportion of their grocery spend is on chips
- Proportion of customers in each customer segment overall to compare against the mix of customers who purchase chips
Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and
plotting the split by these segments to describe which customer segment contribute most to chip sales.
```{r fig.width = 10, fig.align = "center"}
#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER
Total_sales_summary <- data %>% 
  drop_na %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarize(Total_sales = sum(TOT_SALES)) 
ggplot(Total_sales_summary, mapping = aes(x = PREMIUM_CUSTOMER, y = Total_sales)) +
  geom_col() +
  facet_wrap(~LIFESTAGE) +
  labs(title = "Total Sales By Lifestage and Premium Customer")
```
Sales are coming mainly from Budget - older families, Mainstream - young singles/couples, and Mainstream - retirees.
see if the higher sales are due to there being more customers who buy chips.
```{r fig.width = 10, fig.align = "center"}
#### Number of customers by LIFESTAGE and PREMIUM_CUSTOMER
customer_summary <- customerData %>% 
  drop_na %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarize(Number_of_customers = n())
ggplot(customer_summary, mapping = aes(x = PREMIUM_CUSTOMER, y = Number_of_customers)) +
  geom_col() +
  facet_wrap(~LIFESTAGE) +
  labs(title = "Number Of Customers By Lifestage and Premium Customer")
```
There are more Mainstream - young singles/couples and Mainstream - retirees who buy chips. This contributes to there being more sales to these customer segments but this is not a major driver for the Budget - Older families segment. Higher sales may also be driven by more units of chips being bought per customer.
```{r fig.width = 10, fig.align = "center"}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
units_summary <- data %>% 
  drop_na %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarize(avg_units = mean(PROD_QTY)) 
ggplot(units_summary, mapping = aes(x = PREMIUM_CUSTOMER, y = avg_units)) +
  geom_col() +
  facet_wrap(~LIFESTAGE) +
  labs(title = "Average Units Bought By Lifestage and Premium Customer")
```
Older families and young families in general buy more chips per customer
investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.
```{r fig.width = 10, fig.align = "center"}
#### Average price per unit by LIFESTAGE and PREMIUM_CUSTOMER

# Adding price per unit column
data <- mutate(data, UNIT_PRICE = TOT_SALES/PROD_QTY)

# summarizing data and plotting the graph
unit_price_summary <-  data %>% 
  drop_na %>% 
  group_by(LIFESTAGE, PREMIUM_CUSTOMER) %>% 
  summarize(avg_price_per_unit = mean(UNIT_PRICE)) 
ggplot(unit_price_summary, mapping = aes(x = PREMIUM_CUSTOMER, y = avg_price_per_unit)) +
  geom_col() +
  facet_wrap(~LIFESTAGE) +
  labs(title = "Average Price Per Unit Of Chips Bought By Lifestage and Premium Customer")
```
Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than their own consumption.
This is also supported by there being fewer premium midage and young singles and couples buying chips compared to their mainstream counterparts.
As the difference in average price per unit isn't large, check if this difference is statistically different.
```{r}
#### Perform an independent t-test between mainstream vs premium and budget midage and young singles and couples

#filter for mainstream and premium
df_mainstream_premium <- unit_price_summary[unit_price_summary$PREMIUM_CUSTOMER %in% c("Mainstream", "Premium") & unit_price_summary$LIFESTAGE %in% c("MIDAGE SINGLES/COUPLES", "YOUNG SINGLES/COUPLES"),]

#filter for mainstream and budget
df_mainstream_budget <- unit_price_summary[unit_price_summary$PREMIUM_CUSTOMER %in% c("Mainstream", "Budget") & unit_price_summary$LIFESTAGE %in% c("MIDAGE SINGLES/COUPLES", "YOUNG SINGLES/COUPLES"),]

#t-test for mainstream and premium
t_test_mainstream_premium <- t.test(avg_price_per_unit ~ PREMIUM_CUSTOMER,data = df_mainstream_premium)
#t-test for mainstream and budget
t_test_mainstream_budget <- t.test(avg_price_per_unit ~ PREMIUM_CUSTOMER,data = df_mainstream_budget)
t_test_mainstream_premium
t_test_mainstream_budget
```

## Deep dive into specific customer segments for insights
Might want to target customer segments that contribute the most to sales to retain them or further increase sales. Looking at Mainstream - young singles/couples. Do they tend to buy a particular brand of chips.
```{r fig.align = "center"}
#### Deep dive into Mainstream, young singles/couples

data %>% 
  drop_na %>% 
  filter(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream") %>% 
  group_by(BRAND_NAME) %>% 
  summarize(no_of_customers = n()) %>% 
  arrange(-no_of_customers)

```
see that :

* The top three brands patronised by mainstream young single and couples are Kettle, Doritos and Pringles.
* The least patronised brand by mainstream young singles and couples is Cobs.

find out if our target segment tends to buy larger packs of chips.
```{r fig.align = "center"}
#### Preferred pack size compared to the rest of the population
max(data$PACK_SIZE, na.rm = TRUE)
data %>% 
  drop_na %>% 
  filter(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream") %>% 
  group_by(PACK_SIZE) %>% 
  summarize(no_of_purchases = n()) %>% 
  arrange(-no_of_purchases)
```
### INSIGHTS

* It can be seen that the largest pack size is 380.
* The most pack size of chips bought by mainstream young singles and couples is 175 which was purchased 4997 times.
* The largest pack size was bought by mainstream young singles and couples 626 times.
* Therefore, it can be said that mainstream young singles/couples don't tend to buy larger packs of chips.